Effects of epidemiological structure on the transient evolution of HIV virulence

### 2016-12-20 ms/cover_letter/plosHIVcover_resubm2.tex
updating cover letter, tweaking figs; add rev 3 PDF

@@

>>>>

\response{comparing to Shirreff \ldots Figure 1 shows comparison \ldots picking high start virulence, won't see peak at all \ldots very much unknown \ldots implicit assumption that there's not an \emph{interaction} between starting SPVL and characteristics of models \ldots}

@@

\textbf{Reviewer \#3:}
\revcomment{
p. 10 if not counteracted by increased mortality among hosts infected by these reproductive strains.
}

\response{(confusion between transmission rate and  ${\mathcal R}_0$ ---
  clarify in text)}

\revcomment{
p. 11 instead of calling this "epidemiological structure" it may be more appropriate to refer to it as "contact structure". The authors investigated the effect of adding structure to the pattern of sexual contacts. There is nothing intrinsically "epidemiological" about forming (stable and/or once-off) sexual relationships.
}

\revcomment{
p. 11 Given the long time horizons of the simulations (500-4000
years), this simplifying assumption [single-stage model of HIV
progression] may introduce bias, but unfortunately the magnitude and importance thereof is unknown. In more realistic models, one would expect that the high-risk susceptible individuals get depleted at a higher rate than the low risk susceptibles, with the epidemiological effects of this risk gradient playing out over a small number of decades. If or how this dynamic affects the evolution of virulence is unclear but if the goal is explicitly to study this evolution, it is problematic that all models under study are unable to take these demographic effects into account, by design.
}

\revcomment{
p. 12 As noted by reviewer 1, the decision to ignore temporal variation in
infectiousness over the main stages of HIV infection is a serious
limitation of this study. In particular, the epidemiological
contribution of the spike in HIV viral load during acute and early HIV
infection depends on the assumed contact pattern. In models where
individuals are not "trapped" in serially monogamous relationships,
recently infected individuals can make efficient use of this phase as
they can transmit to new or additional sexual partners as soon as they
are infected. In contrast, in models that only allow for stable
partnerships, this spike in infectiousness will "go to waste" as the
newly infected individuals are most likely still in the relationship
with the partner that they acquired the infection from during these
initial few months after HIV acquisition. For highly virulent viral
strains (high SPVL and short survival) such window of opportunity is
more important than for viral strains that allow the host to transmit
for a decade or more.
}

\response{make sure we have a caveat/comment in text}

\revcomment{
p. 12 Showing that the model with only pair formation dynamics (~
Shirreff et al.) produces qualitatively similar results to those from
the Shirreff model, does not prove or suggest that the effect of
adding additional extra-couple transmission is robust against the
model's structure with respect to the stages of HIV infection and
associated infectiousness levels.
}

\response{same thing again}

\revcomment{
p. 21 The authors touch on two weaknesses of their analysis here: 1. The level of sufficient realism is questionable, despite great efforts to take into account various sources of variation (with several thousands of compartments as a result).
2. The real-life implications of their main metrics (time to reach
peak virulence, time to reach equilibrium virulence, progression time
to AIDS at peak virulence) is unclear, because in all model scenarios
(strain-specific) transmission and disease progression/mortality
parameters stay constant over time. However, dramatic increases in the
uptake of HIV treatment (ART) have been observed across the world, and
trends in sexual risk behaviour may have changed both in the pre-ART
and ART era.
}

\revcomment{
p. 22  The model with heterogeneous levels of sexual activity (the
"heterogeneity" model) is an extension of the pair formation + epc
model. However, in principle, this extra layer of heterogeneity in
sexual contact rate could have been added to the other 5 "base" models
as well. In the current comparison it is not possible to assess the
relative effect of adding extra-couple and uncouple transmission
versus contact heterogeneity.
}

\revcomment{
p. 23 
This sentence [``These differences are practically as well as
scientifically important''] seems to belong better in the discussion section. Here, the authors do not explain what the practical or scientific importance of the differences is.
}

\revcomment{
p. 27 The main finding of this analysis seems to be that allowing "unstructured" transmission events, be it extra-couple or uncoupled, influences the expected trajectory of SPVL evolution over rather long time scales. As Reviewer 1 notes, this is indeed a rather theoretical finding. In my option, two key questions remain unaddressed: 1. Why should we care more about the effect of this aspect of the assumed contact pattern than the effects of other aspects of the contact pattern and disease process (e.g. age-mixing, stage-varying infectiousness, ART status, depletion of high-risk susceptibles)?
2. What are the practical implications of these findings for HIV
epidemiology? For HIV prevention and treatment? Does it help us in any
way to interpret empirical time-trends in SPVL or disease progression?
}
